Fine-tuning super-resolution for real-world aeromagnetic surveys

### 2024-06-06 intro.tex
Address team comments

\section{Magnetic potential fields}
The Earth has a naturally occuring magnetic force field, approximating a dipole oriented toward the top and bottom of the planet.
Several properties of the magnetic field are important to the machine learning methods in this thesis.
First, that a field refers to a set of functions of space and time, and a vector field such as the magnetic field has three orthogonal components, with a function for each component.
Neural networks with sufficient parameters are recognised as universal function approximators \parencite{hornikMultilayerFeedforwardNetworks1989}, and should therefore be capable of approximating these functions of the magnetic field.
Second, the magnetic field has a force function with continuous derivatives, and is conservative \parencite{kelloggFoundationsPotentialTheory1967}.
The force function is known as the potential field, and the magnetic field is the gradient of the potential field.
This property is important later in this thesis, when calculating spatial derivatives and considering constraining neural networks with physics informed loss functions.

The magnitude of the magnetic field is the total magnetic intensity (TMI) \parencite{blakelyPotentialTheoryGravity1996}, with the unit nanotesla (nT).
Magnetic surveys in exploration geophysics generally create TMI products, and it may be measured directly or calculated from measurements of each orthogonal component.
To support various scientific endeavours, a model of this field is periodically constructed --- the International Geomagnetic Reference Field (IGRF).
The residual between the IGRF and measured TMI at a specific time and location is termed magnetic anomaly.
It may be negative, and can exceed several thousand nanotesla in the Earth's environment.
One of the key petrophysical properties influencing the magnetic anomaly is magnetic susceptibility, itself predominantly influenced by magnetite, a mineral commonly present across geological settings \parencite{clarkNotesRockMagnetization1991}.
Rocks may also be intrinsicly magnetised with remnant magnetisation.
Because of these known petrophysical variations in rock magnetisation, stratigraphies can be differentiated and delineated in magnetic maps, and their composition interpreted.

@@

\Cref{ch:paper3} reports the use of coordinate multilayer perceptron (CMLP) neural networks for implicit neural representation (INR) of a surveyed potential field extent.
With the proposed method, scattered line data can be encoded into a neural network model, and grids created by querying cells at regular coordinates.
% CMLP networks are highly suited for processing spatial data such as point sampled geophysics surveys.
Additionally, spatial derivatives are calculated on the learned function directly.
%TODO ensure refer back to aims
Both tasks are fundamental to interpretation and processing in geophysics, and by commencing the methods from point sample data, the number of individual processing steps is reduced, and the application is fully trainable.
To our knowledge, this research is the first of its kind in geophysical literature.

>>>>

\section{Magnetic potential fields}
The Earth has a naturally occuring magnetic field, approximating a dipole oriented toward the top and bottom of the planet.
Several properties of the magnetic field are important to the machine learning methods in this thesis.
First, that a field refers to a set of functions of space and time.
The magnetic field is a vector field, with three orthogonal components.
Neural networks with sufficient parameters are recognised as universal function approximators \parencite{hornikMultilayerFeedforwardNetworks1989}, and should therefore be capable of approximating the functions of the magnetic field.
Second, the curl of the magnetic vector field is a force function known as the magnetic potential field \parencite{kelloggFoundationsPotentialTheory1967}.
The magnitude of the magnetic vector field is the total magnetic intensity (TMI) \parencite{blakelyPotentialTheoryGravity1996}, with the unit nanotesla (nT).
Magnetic surveys in exploration geophysics generally create TMI products, where TMI may be measured directly or calculated from measurements of each vector component.
To support various scientific endeavours, a model of this field is periodically constructed --- the International Geomagnetic Reference Field (IGRF).
The residual between the IGRF and measured TMI at a specific time and location is termed magnetic anomaly.
It may be negative, and can exceed several thousand nanotesla in the Earth's environment.
One of the key petrophysical properties influencing the magnetic anomaly is magnetic susceptibility, itself predominantly influenced by magnetite, a mineral commonly present across geological settings \parencite{clarkNotesRockMagnetization1991}.
Rocks may also be intrinsicly magnetised with remnant magnetisation.
Because of these known petrophysical variations in rock magnetisation, surveys of the magnetic field can delineate and differentiate stratigraphic units and infer their composition.

@@

\Cref{ch:paper3} reports the use of coordinate multilayer perceptron (CMLP) neural networks for implicit neural representation (INR) of a surveyed potential field extent.
With the proposed method, scattered line data can be encoded into a neural network model, and grids are created by querying cells at regular coordinates.
% CMLP networks are highly suited for processing spatial data such as point sampled geophysics surveys.
Additionally, spatial derivatives are calculated on the learned function directly.
%TODO ensure refer back to aims
Both tasks are fundamental to interpretation and processing in geophysics, and by commencing the methods from point sample data, the number of individual processing steps is reduced, and the application is fully trainable.
To our knowledge, this research is the first of its kind in geophysical literature.

### 2024-06-06 paper2.tex
Adjust chapter titles
Prompted by reviewer 2 to better present the thesis
Address team comments

\title{Super-resolution of Aeromagnetic Survey Data}
\author[1*]{Luke Thomas Smith}
\author[1]{Tom Horrocks}
\author[1]{Eun-Jung Holden}
\author[1]{Daniel Wedge}
\author[2]{Naveed Akhtar}
\affil{Centre for Data-Driven Geoscience, School of Earth Sciences, The University of Western Australia (M006), 35 Stirling Highway, 6009 Perth, Australia}
\affil{Department of Computer Science and Software Engineering, The University of Western Australia (M002), 35 Stirling Highway, 6009 Perth, Australia}
\date{\today}
% \maketitle{}

@@

In order to improve the generalisation of an example-based neural network for regions without high-resolution coverage, we propose using transfer learning \parencite{tanSurveyDeepTransfer2018} to fine-tune a synthetic data baseline model with high-resolution, real-world survey grid data.
The synthetic geophysical data \parencite{jessellNoddyverseMassiveData2022} are forward modelled from 3D petrographic voxel models with varied geological histories.
These provide a diverse set of features, but these lack truly realistic characteristics, such as high frequency near-surface magnetisation, and features from complex geological processes such as metasomatism or remnant magnetisation.

@@

Instead of increasing the sampling density, the detail in potential field grids can be enhanced by calculating the potential as if it were sampled at a smaller source-sensor distance, in a process known as downward continuation \parencite{bullardDeterminationMassesNecessary1948}.
At large source-sensor distances, high frequencies are diminished below the noise floor to the point of unobservability, resulting in grids that show only broad scale features, typically representing spatially extensive features deep in the crust.
Downward continuation can be performed with the grid representation of the field, or by modelling causative sources and forward modelling the response at another height \parencite{pilkingtonPotentialFieldContinuation2017}.
In either approach, it is common for noise in the sampled grid product to be significantly increased in the downward continued product, limiting the useful vertical continuation extent of of the method to six times the cell size \parencite{dampneyEquivalentSourceTechnique1969,zuoDownwardContinuationTransformation2020}.
Resolution enhancement through neural network approaches to downward continuation \parencite{liStableDownwardContinuation2023,yeHighprecisionDownwardContinuation2022} is a promising research direction in super-resolution geophysics, but the training data are distinct.
The proposed line-spacing based methods use data prepared using high- and low-resolution survey designs, while neural network continuation approaches prepare data using the above numerical continuation methods.

@@

Deep learning techniques for SR typically use paired high- and low-resolution image data, with the objective to learn the transform from low resolution to high.
This is an ill posed challenge, because many high-resolution images share the same low-resolution representation \parencite{dongImageSuperresolutionUsing2016}.
For photographs and natural image rasters with large representative training corpuses available, such as ImageNet \parencite{dengImageNetLargescaleHierarchical2009} and DIV2K \parencite{agustssonNTIRE2017Challenge2017}, SR methods have progressed to the point that high quality enhancement at four times scale is readily accessible to end-users \parencite[e.g.][]{wangRealESRGANTrainingRealWorld2021}.
The performance of SR for photographs and digital images has been intensely explored \parencite[see a review in][]{moserHitchhikerGuideSuperResolution2023}.
Prior work specific to this study are reviewed in \Cref{sec:2nnsr}.

>>>>

\title{Fine-tuning super-resolution for real-world aeromagnetic surveys}
\author[1*]{Luke Thomas Smith}
\author[1]{Tom Horrocks}
\author[1]{Eun-Jung Holden}
\author[1]{Daniel Wedge}
\author[2]{Naveed Akhtar}
\affil{Centre for Data-Driven Geoscience, School of Earth Sciences, The University of Western Australia (M006), 35 Stirling Highway, 6009 Perth, Australia}
\affil{Department of Computer Science and Software Engineering, The University of Western Australia (M002), 35 Stirling Highway, 6009 Perth, Australia}
\date{\today}
% \maketitle{}

@@

In order to improve the generalisation of an example-based neural network for regions without high-resolution coverage, we propose using transfer learning \parencite{tanSurveyDeepTransfer2018} to fine-tune a synthetic data baseline model with high-resolution real-world survey grid data.
The synthetic geophysical data \parencite{jessellNoddyverseMassiveData2022} are forward modelled from 3D petrographic voxel models with varied geological histories.
These provide a diverse set of features, but these lack truly realistic characteristics, such as high frequency near-surface magnetisation, and features from complex geological processes such as metasomatism or remnant magnetisation.

@@

Instead of increasing the sampling density, the detail in potential field grids can be enhanced by calculating the potential as if it were sampled at a smaller source-sensor distance, in a process known as downward continuation \parencite{bullardDeterminationMassesNecessary1948}.
At large source-sensor distances, high frequencies are diminished below the noise floor to the point of unobservability, resulting in grids that show only broad scale features, typically representing spatially extensive features deep in the crust.
Downward continuation can be performed with the grid representation of the field, or by modelling causative sources and forward modelling the response at another height \parencite{pilkingtonPotentialFieldContinuation2017}.
In either approach, it is common for noise in the sampled grid product to be significantly increased in the downward continued product, limiting the useful vertical continuation extent of of the method to six times the cell size \parencite{dampneyEquivalentSourceTechnique1969,zuoDownwardContinuationTransformation2020}.
Resolution enhancement through neural network approaches to downward continuation \parencite{liStableDownwardContinuation2023,yeHighprecisionDownwardContinuation2022} is a promising research direction in super-resolution geophysics, but the problem formulation is distinct.
The proposed method attempts to predict missing high frequencies with example based learning from similar geologies, while the prior methods attempt to vertically continue grid data using deep learning methods.

@@

Deep learning techniques for SR typically use paired high- and low-resolution image data, with the objective to learn the transform from low resolution to high.
This is an ill posed challenge, because many high-resolution images share the same low-resolution representation \parencite{dongImageSuperresolutionUsing2016}.
For photographs and natural image rasters with large representative training corpuses available, such as ImageNet \parencite{dengImageNetLargescaleHierarchical2009} and DIV2K \parencite{agustssonNTIRE2017Challenge2017}, SR methods have progressed to the point that high quality enhancement at four times scale is readily accessible to end-users \parencite[e.g.][]{wangRealESRGANTrainingRealWorld2021}.
The performance of SR for photographs and digital images has been extensively explored \parencite[see a review in][]{moserHitchhikerGuideSuperResolution2023}.
Prior work specific to this study are reviewed in \Cref{sec:2nnsr}.
